A dual role for PGLYRP1 in host defense and immune regulation during B. pertussis infection

Version 1

Abstract

Bordetella pertussis, the etiologic agent of whooping cough, remains a serious public health concern despite widespread vaccination. Improved therapeutics and vaccines are urgently needed to treat and prevent pertussis disease. Host recognition of bacterial peptidoglycan (PGN), including B. pertussis extracellular PGN fragment tracheal cytotoxin (TCT), shapes the immune response to infection. Peptidoglycan recognition proteins (PGLYRPs) are a conserved family of innate immune molecules which bind bacterial PGN. While they function as immune signaling receptors in arthropods, PGLYRPs in mammals have thus far been primarily recognized for their bactericidal activity. Previously thought to function only as antimicrobial peptides in mammals, the immune modulatory roles of this family of peptidoglycan recognition proteins are beginning to gain greater appreciation. Peptidoglycan recognition protein 1 (PGLYRP1) is a secreted antimicrobial protein. However, its role in mammalian host defenses and immune signaling during infection with Gram-negative pathogens, such as B. pertussis, remain largely unknown. Here, we identify a dual role for PGLYRP1 in modulating host immune responses to B. pertussis. Using knockout mice, single-cell and bulk transcriptomics and functional assays, we show that PGLYRP1 has bactericidal activity against B. pertussis in vitro and promotes early bacterial control in vivo. PGLYRP1 also dampens inflammatory responses and impedes bacterial killing later in infection. Mechanistically, PGLYRP1 enhances nucleotide oligomerization domain (NOD)-1 signaling in response to TCT while suppressing NOD2- and triggering receptor expressed on myeloid cells-1 (TREM-1)-mediated inflammatory pathways. TCT-bound PGLYRP1 selectively impairs TREM-1 activation compared to PGNs from other bacteria, revealing a novel bacterial immune evasion strategy. These findings demonstrate that B. pertussis co-opts PGLYRP1 to temper inflammation and alter immune signaling, revealing a novel immune evasion mechanism of manipulating the availability and structure of their exogenous peptidoglycan, revealing implications for host-pathogen evolution, vaccine design and host-directed therapeutics.

Introduction

Bordetella pertussis is a highly contagious respiratory pathogen and the causative agent of whooping cough, a disease characterized by severe, prolonged coughing and airway inflammation1-3. Despite widespread vaccination efforts, B. pertussis remains a public health concern due to waning vaccine-induced immunity and the pathogen’s capacity to evade host immune defenses4. Host detection of microbial components, including peptidoglycan fragments, is essential for initiating immune responses to infection. Understanding how these host-pathogen interactions shape disease outcomes is critical for developing improved strategies to fight infections.

Peptidoglycan (PGN) is a fundamental component of the bacterial cell wall, composed of repeating disaccharide chains cross-linked by short peptides5. PGN structure is an important determinant of innate immune recognition and responses6, 7. During cell wall turnover, some pathogenic bacterial species, including B. pertussis, release PGN fragments into the extracellular environment8-10. The biological roles of these fragments in infection and immunity are poorly defined.

B. pertussis releases a disaccharide-tetrapeptide monomeric PGN fragment known as tracheal cytotoxin (TCT)11, 12. TCT differs from B. pertussis cell wall PGN by having a 1,6-anhydro group on its MurNAc sugar due to incomplete PGN turnover11. The functional consequences of this 1,6-anhydro MurNAc modification are poorly understood13. TCT is recognized by the cytosolic pattern recognition receptor (PRR) NOD1 to activate NFkB-driven immune responses14, 15. In vitro and ex vivo studies suggest that recognition of TCT by NOD1 in non-ciliated mucus-secreting cells leads to the accumulation of nitric oxide within neighboring ciliated airway epithelial cells and subsequent ciliostasis and extrusion of ciliated cells16, 17. We recently demonstrated that TCT has broader immunomodulatory effects, skewing PGN-sensing toward NOD1 and away from NOD2, thereby dampening a stronger NOD2-dependent, pro-inflammatory cytokine production and generation of immune memory18. However, the mechanisms by which TCT dampens inflammation and the broader immunological consequences of its release during infection, especially in the context of PGN-sensing by PRRs remain poorly defined.

Peptidoglycan Recognition Proteins (PGLYRPs) are a conserved family of soluble extracellular innate immune receptors that bind PGN19. In arthropods, PGLYRPs act as extracellular pattern recognition receptors (PRR) that trigger downstream immune pathways20. In mammals, they have been characterized as bactericidal proteins that target PGN on Gram-positive bacterial surfaces21, 22. However, recent studies have expanded their role beyond antimicrobial defense, suggesting function in regulating host immune responses. PGLYRP1 is a secreted protein stored in neutrophil granules, that is directly bactericidal against several bacterial species23. Recent work suggests that mammalian PGLYRPs also regulate adaptive immunity, including CD8+ T cell responses, in autoimmunity and cancer, and act as an accessory protein for the activation of inflammation amplifying receptor TREM-1 and NOD2-dependent signaling24-26.

TREM-1 is a PRR which amplifies inflammatory responses during infection27. Activation of TREM-1 by its ligands, including PGLYRP1-PGN complexes, enhances production of proinflammatory cytokines and chemokines28. Excessive TREM-1 activation is associated with inflammatory pathology in sepsis and other diseases29. Despite growing interest in PGLYRPs, their roles in host-pathogen interactions, particularly in the context of Gram-negative bacteria and secreted PGN fragments, remains poorly defined.

Here, we demonstrate that PGLYRP1 plays a dual role in host responses to B. pertussis infection, promoting early bactericidal responses and fine-tuning inflammation in response to distinct PGN structures. We demonstrate that exogenous PGLYRP1 enhances NOD1 activation by TCT but failed to enhance muramyl dipeptide (MDP) activation of NOD2. Further, PGLYRP1 directly activates the pro-inflammatory receptor TREM-1, a response that is potentiated by both Lys- and DAP-type PGN from Gram-positive bacteria Staphylococcus aureus or Bacillus subtilis. PGLYRP1-mediated activation of TREM-1 was unexpectedly suppressed by B. pertussis TCT. These findings reveal that PGLYRP1 is not just a bactericidal protein, but also a context-specific immunomodulator whose function is shaped by the structure of the PGN it encounters.

This work uncovers a previously unrecognized mechanism by which extracellular PGN and innate receptors together shape host inflammation during infection in a structure-specific manner.

Results

Opposing roles of PGLYRP1 in early and late host defenses to B. pertussis

PGLYRP1 has known bactericidal activity in vitro, particularly against Gram-positive bacteria23. To investigate whether PGLYRP1 contributes to host antibacterial defenses against B. pertussis, we intranasally infected adult BALB/c wild-type (WT) and PGLYRP1-knockout (PGLYRP1 KO) mice and assessed lung bacterial burden at 4- and 7-days post infection (DPI). Surprisingly, PGLYRP1 had opposing impacts on bacterial burden at these time points (Fig. 1A&B). At 4DPI, PGYLRP1 KO mice exhibited significantly higher bacterial burdens compared to WT controls (p < 0.01), indicating that PGLYRP1 contributes to early bacterial control. However, by 7DPI PGLYRP1 KO mice had significantly lower bacterial burden compared to WT mice (p < 0.001), suggesting that PGLYRP1 impairs bacterial clearance later in infection, potentially by modulating host immune responses.

PGLYRP1 has a dual role in host defense against Bordetella pertussis.

Lung bacterial burdens in wild-type BALB/c (WT, black) and PGLYRP1-knockout (PGLYRP1 KO, red) at 4-(A) and 7-(B) days post-infection. Violin-plot visualizing single-cell expression levels of Pglyrp1 across cell types in the lungs of C57/B6 at 4DPI or sham challenged mice, assessed via analysis of single-cell RNA sequencing data (C). RNAscope in situ hybridization visualization of Pglyrp1 transcripts in FFPE sections of B. pertussis infected mouse lung tissue (D). Each dot represents a single Pglyrp1 mRNA molecule. Images are presented at 20X and 10X magnification Ex vivo bone marrow derived neutrophil killing assay (E). Neutrophils were isolated from BALB/c (WT) or PGLYRP1 KO mice and incubated with B. pertussis for 2 (circles) or 24 (square) hours and CFU enumerated by serial dilution on Bordet Gengou agar. In vitro bacterial killing assay assessing recombinant murine PGLYRP1 (mPGLYRP1, red) bactericidal activity at 6- and 24-hours post-incubation compared to BSA control against WT (F) or a TCT non-releasing strain (TCT-) (G). Enumeration of CFU following incubation of mPGLYRP1 with wild-type B. pertussis (parental) or mutant lacking the BpsB gene of the Bordetella polysaccharide (Bps) operon (H). Data represents CFU as a percentage of starting inoculum recovered at indicated timepoint as assessed by serial dilution of culture on BG agar. Data are presented as mean ± SEM; significance determined by unpaired t-test or ANOVA as appropriate. **, p-value < 0.01, *** p-value < 0.005, **** p-value < 0.001

To elucidate the potential impact of PGLYRP1 on host immune responses, we first performed single cell RNA sequencing on lungs from uninfected and B. pertussis-infected C57BL/6 mice at 4DPI to identify PGLYRP1 expressing cell-types following infection. In uninfected lungs PGLYRP1 expression was most prominent in neutrophils (Fig. 1C). Upon infection the PGLYRP1-expressing cell-types expanded to include ciliated cells, secretory cells, inflammatory monocytes and lymphocytes in addition to neutrophils (Fig. 1C). RNAscope in situ hybridization confirmed expression in the airways of infected mice (Fig. 1D).

To determine whether PGLYRP1 has bactericidal activity against B. pertussis, we performed in vitro and ex vivo killing assays. Based on transcriptomic data (Fig. 1C) we assessed the bactericidal potential of BMDNs from WT and PGLYRP1 KO mice incubated with B. pertussis for 2 or 24 hours. At 24 hours, PGLYRP1 KO neutrophils failed to control viable bacteria compared to WT neutrophils (Fig. 1E), indicating that PGLYRP1 contributes to neutrophil mediated killing of B. pertussis. In a cell-free assay, recombinant murine PGLYRP1 (mPGLYRP1) significantly reduced bacterial CFU after 24 hours of incubation compared to BSA control (Figure 1F, p<0.001), confirming its bactericidal potential.

TCT release by B. pertussis is a product of inefficient recycling by the permease AmpG11. Replacing B. pertussis AmpG, with E. coli AmpG resulted in a strain (TCT-) which releases 99% less TCT (TCT(-)), while deletion of AmpG created a strain with 24-fold greater TCT release (TCT(+))30. Since PGLYRP1 binds bacterial PGN, we hypothesized that B. pertussis TCT, an extracellular PGN fragment, may act as a decoy to divert PGLYRP1 activity. However, no increase in the bactericidal activity of recombinant mPGLYRP1 was noted between WT or TCT-deficient strains following incubation with PGLYRP1 (Fig. 1G), arguing against this hypothesis. Given prior work showing that Bordetella polysaccharides confer resistance to antimicrobial peptides31, we tested whether extracellular polysaccharides inhibit PGLYRP1 function. The Bps operon is required for effective production of B. pertussis surface polysaccharide32. Using a BpsB-deficient strain we observed significantly enhanced killing by mPGLYRP1 compared to WT bacteria (Fig. 1H, p< 0.01), suggesting that endogenous polysaccharides shield B. pertussis from PGLYRP-mediated killing.

Together, these results indicate that PGLYRP1 contributes to both bacterial killing and regulation of host immune responses during B. pertussis infection. PGLYRP1 enhances early bacterial clearance via neutrophil-mediated and direct bactericidal mechanisms but may suppress immune-mediated clearance later in infection. While TCT does not appear to interfere with PGLYRP1 bactericidal activity, Bordetella polysaccharides, previously associated with biofilm production32, do, highlighting a key bacterial evasion strategy against host innate defenses.

Peptidoglycan fragment TCT enhances PGLYRP1-mediated suppression of host responses in pertussis

Given the opposing effects of PGLYRP1 on bacterial burden at 4DPI and 7DPI, we hypothesized that PGLYRP1 both exerts bactericidal activity while also suppressing protective immune responses, indirectly impairing bacterial killing. Recently, PGLYRP1 has been associated with PGN recognition by innate immune receptors, including NOD2 and TREM-124, 26. However, B. pertussis releases a NOD1 activating PGN fragment, TCT14.

To assess the role of PGLYRP1 in B, pertussis-driven inflammation, we performed histological analysis of lungs from WT and PGLYRP1 KO mice at 4 and 7DPI. We observed significantly reduced immunopathology, based on the percentage and severity of bronchovascular bundle formation and infiltration of immune cells into alveolar spaces, in PGLYRP1 KO mice at both timepoints (Fig. 2A, p < 0.05 at 4DPI, p < 0.01 at 7DPI). These findings suggest that PGLYRP1 contributes to inflammatory lung damage during B. pertussis infection.

Assessment of PGLYRP1’s role in pulmonary inflammation and transcriptional responses during Bordetella pertussis infection.

Semi-quantitative scoring of hematoxylin and eosin (H&E) stained lung sections from wild-type BALB/c (black dots) and PGLYRP1 knockout (red dots) mice at 4- and 7-days post-infection with B. pertussis (A). Data represents average scores of 3 blinded investigators scoring degree and percentage of peribronchial infiltration and alveolar consolidation. Bulk RNA sequencing was performed on lung homogenates collected from WT and PGLYRP1 KO mice at 4DPI to assess differential gene expression. Volcano plots (B,D) and heat-maps (C,E) showing total (B,D) and immune-related (C,E) differentially expressed genes between B. pertussis challenged BALB/c (WT) and PGLYRP1 KO mice (B,C) or PGLYRP1 KO mice challenged with parental wild-type B. pertussis (WT) or TCT-under releasing strain TCT(-). (F) KEGG enrichment pathway analysis was performed on differentially expressed genes from lung tissue isolated from mice challenged with WT or TCT(-) B. pertussis. Data represent the mean of biological replicates; statistical analyses were performed using adjusted p-values as indicated in the main text. *, p-value < 0.05, **, p-value < 0.01

To investigate PGLYRP1s role in immune signaling during B. pertussis infection, we performed bulk RNA sequencing on lungs from infected BALB/c (WT) and PGLYRP1 KO mice. PGLYRP1 KO mice exhibited increased expression of pro-inflammatory cytokines (e.g., IL6, IL23A, IL1A, IL1B, and IL36G) and chemokines (e.g., CCL2, CCL3, CCL12, and CCL28) alongside reduced expression of anti-inflammatory cytokine IL10 (Fig. 2B-C). This transcriptomic profile suggests a model in which PGLYRP1 dampens key cytokine responses while paradoxically promoting immunopathology.

PGLYRP1 is a soluble PGN-binding protein. B. pertussis releases an extracellular PGN fragment, tracheal cytotoxin (TCT). We have recently demonstrated that TCT inhibits immune responses to B. pertussis18.We hypothesized that interactions with TCT contribute to the influence of PGLYRP1 on host inflammatory responses to B. pertussis. To test this, we compared the transcriptional responses of PGLYRP1 KO mice challenged with TCT-producing (parental) or TCT-deficient B. pertussis strains. Both strains induced a similar pattern of cytokine upregulation in the absence of PGLYRP1, but this increased inflammation was more robust in the absence of TCT (Fig. 2D&E). Specifically, PGLYRP1 KO mice infected with TCT(-) bacteria exhibited greater expression of IL23A, IL6, IL1B, and IL36G compared to WT infection, suggesting that TCT may potentiate the immune-dampening activity of PGLYRP1.

PGLYRP1 has previously been shown to stimulate inflammatory responses to PGN26. These findings indicate that the immunomodulatory function of PGLYRP1 is informed by the PGN it interacts with. TCT release may be a strategy employed by B. pertussis to co-opt PGLYRP1 function and suppress protective inflammation, thereby promoting pathogen persistence. Supporting this, KEGG pathway analysis on differential gene expression of 4DPI WT (parental) and TCT(-) lungs confirmed enrichment of cytokine–cytokine receptor interactions, NOD signaling, and IL-17/TNF pathways in the presence of TCT (Fig. 2F).

TCT-PGLYRP1 interactions drive NOD1 signaling

To understand how TCT modulates PGLYRP1-dependent immune signaling, we investigated known mechanisms by which PGLYRP1 influences host PRR signaling. Recent studies suggest that intracellular PGLYRP1 can act as a co-factor in NOD2 signaling24 and TCT is a NOD1-exclusive agonist. Hence, we hypothesized that PGLYRP1 may enhance NOD1 signaling in response to the muropeptide TCT. To test this, recombinant PGLYRP1 was incubated with TCT before stimulation of NOD1 and NOD2 reporter cells.

PGLYRP1 alone did not induce NOD1 activation (Fig. 3A). TCT alone robustly activated NOD1 (Fig. 3A). Notably, both murine (Fig. 3A, p < 0.005) and human (Fig. 3B, p < 0.001) PGLYRP1 significantly enhanced TCT-mediated NOD1 activation compared to TCT alone. This enhanced activation was specific to NOD1 as neither TCT nor TCT+PGLYRP1 activated NOD2 reporter cells (data not shown) since TCT is not a NOD2 agonist. On the other hand, mPGLYRP1 decreased NOD2 recognition of NOD2 agonist muramyl dipeptdide (MDP, p < 0.01) (Fig. 3C), suggesting exogenous PGLYRP1 boosts responses to NOD1 stimulating PGN structures like TCT but dampens responses to NOD2 stimulatory PGNs.

PGLYRP1 selectively modulates NOD1 and NOD2 signaling in response to Bordetella pertussis peptidoglycan.

(A–B) HEK293 reporter cells expressing either mouse (A) or human (B-D) NOD1 or NOD2 were stimulated with recombinant mouse (A) or human (B-D) PGLYRP1 (12.5 ug/mL), purified tracheal cytotoxin (TCT) (A,B), muramyl dipeptide (MDP, NOD2 agonist) (C) or conditioned bacterial growth media (OD = 0.8) (D). NF-κB-driven SEAP reporter activity was measured to assess pathway activation. Reporter activation was quantified after 18–24 hours. Single-cell RNA sequencing was performed on lung tissue from BALB/c mice infected with B. pertussis at 4 days post-infection. Neutrophils were subjected to further sub-clustering and visualized as a UMAP (E). Violin plots visualizing expression of Nod1, Nod2, IL1a and Pglyrp1 across neutrophil sub-clusters (F). Volcano plot demonstrating differentially expressed genes in Nod1 (blue) or Nod2 (red) expressing neutrophils (G) Single-cell analyses represent combined data from two mice per group. Statistical analyses were performed using Student’s t-test or adjusted p-values as appropriate. **, p-value < 0.01, *** p-value < 0.005, **** p-value < 0.001

To determine if PGLYRP1 modulates NOD responses to other potential PGNs released by B. pertussis, we stimulated reporter cells with mPGLYRP1 incubated with conditioned media from WT or TCT(-), or TCT(+) cultures. Consistent with our results using purified TCT, mPGLYRP1 amplified NOD1 signaling in response to B. pertussis conditioned media, likely due to increased recognition of extracellular TCT (Fig. 3D, p < 0.01). In contrast, NOD2 signaling was significantly suppressed by mPGLYRP1, despite the presence of unknown non-TCT muropeptides potentially capable of activating NOD2 (Fig. 3D), consistent with our results with purified PGN (Fig. 3A-C). Together, these findings suggest that exogenous PGLYRP1 acts as a selective modulator of NOD signaling, enhancing NOD1 activation while dampening NOD2 responses.

To assess the functional consequence of skewed NOD1 versus NOD2 expression and signaling during infection, we analyzed single-cell RNA sequencing data from lungs of BALB/c mice at 4DPI. We examined NOD1 and NOD2 expression across neutrophil sub-clusters (Fig. 3E&F), as neutrophils are the primary expressors of PGLYRP1 during B. pertussis infection (Fig. 1C). NOD2-expressing neutrophils showed higher expression of pro-inflammatory genes including Il1a, Ccl3 and Ptgs2 (COX-2) (Fig. 3F&G). In contrast, NOD1 expression correlated with genes involved in anti-inflammatory immune responses: Pla2g7 (degrades platelet activating factor), Tmsb4x which inhibits neutrophil migration, promotes repair and inhibits NFkB and increased expression of PGLYRP1, which may further positively regulate these pathways. This identifies a correlation between pro-inflammatory cytokine expression and NOD2 signaling in neutrophils. These results highlight the heightened inflammatory phenotype of NOD2+ neutrophils and suggest that TCT-mediated skewing of NOD responses towards NOD1, amplified by PGLYRP1, may be a mechanism to suppress inflammation and promote B. pertussis persistence.

TCT-PGLYRP1 complexes dampen TREM-1 activation, limiting inflammation during B. pertussis infection

In addition to their role in NOD signaling, PGLYRP1 complexes have been shown to activate TREM-1, driving NFkB-mediated inflammatory responses26. This ability to activate TREM-1 is boosted by the presence of a DAP-type PGN isolated from E. coli26. To first determine if TREM-1 contributes to host responses to B. pertussis, we first assessed TREM-1 expressing cell-types following B. pertussis infection in mice using single cell RNA-sequencing. Following B. pertussis infection, TREM-1 was primarily expressed by neutrophils, with lesser expression noted in monocytes and macrophages (Fig. 4A).

TCT inhibits PGLYRP1 signaling via TREM-1

Violin plot generated from single-cell RNA sequencing (scRNA-seq) displaying TREM-1 expression on neutrophils (green), alveolar macrophages (AM, gold), and inflammatory monocytes (IM) at 4DP1 (A). TREM-1 activation measured via NFAT-driven luminescence (relative luminescence units, RLU) in TREM-1 reporter cells treated with recombinant human PGLYRP1 with or without 5μg/mL of Lys-type peptidoglycan derived from Staphylococcus aureus, tracheal cytotoxin (TCT) or a DAP-type PGN from Bacillus subtilis (5 μg/mL) (B). Diagram depicting proposed model of exogenous PGLYRP1 boosting NOD1 and TREM-1 signaling, but dampening NOD2 signaling. Diagram further depicts B. pertussis co-opting these pathways via release of TCT to block TREM-1 signaling and bias NOD signaling (C). All TREM-1 activation assays were performed with biological triplicates (n = 3). Statistical analyses were performed using unpaired two-tailed Student’s t-test. *, p-value < 0.05, **, p-value < 0.01, *** p-value < 0.005, **** p-value < 0.001

To determine if TCT influences PGLYRP1-dependent TREM-1 activation, we assessed the ability of TCT, a 1,6-anhydrous DAP-type PGN monomer, to enhance PGLYRP1-mediated TREM-1 activation compared to PGN isolated from S. aureus, a non 1,6-anhydrous Lys type PGN, and B. subtilis, non 1,6-anhydrous DAP-type PGN monomer. Given the anti-inflammatory role of TCT observed in our data (18 and Figs 2 and 3), we proposed that TCT may engage PGLYRP in a manner that fails to promote, or actively impairs, TREM-1 activation compared to other PGN structures. Using TREM-1 reporter cells, we noted activation of TREM-1 by hPGLYRP1, which was significantly enhanced in the presence of PGN from S. aureus (SA) or Bacillus subtilis (BS (Fig. 4B, p < 0.01), confirming that PGLYRP1 agonism of TREM1 is enhanced by the presence of these species’ PGN structures26. TCT which contains a DAP-type peptide tail, but a 1,6-anhydro bond containing glycan moiety fails to potentiate TREM-1 signaling (Fig. 4C). Instead, we observed a significant reduction of TREM1 activation compared to PGLYRP1 alone, suggesting that TCT antagonizes PGLYRP1’s ability to activate TREM-1, opposite the effect of PGNs. Previous work from our group suggests that TREM-1 signaling exacerbates inflammation in response to B. pertussis infection as treatment with TREM-1 inhibitors significantly decreased pulmonary inflammation during B. pertussis infection33. These findings highlight the importance of bacterial regulation of TREM-1 in B. pertussis pathogenesis and disease and the potential impact of TCT-mediated skewing of TREM-1 responses.

Discussion

Our study reveals a previously unrecognized role for the PRR, PGLYRP1, in orchestrating both antibacterial defenses and inflammatory responses during Bordetella pertussis infection. While traditionally characterized as a bactericidal effector which mediates its effects by binding PGN on the surface of Gram-positive bacteria, emerging work suggests its role may extend to mammalian immune signaling25. The contribution of PGLYRP1 to PRR mediated inflammatory responses in mammals and in particular how PGN structure impacts the outcomes of these interactions, remains largely unexplored. Our findings demonstrate that PGLYRP1 functions as a PGN structure-specific interpreter which translates structure into context and pathogen specific immune signaling. This dual role expands our understanding of how innate immune receptors fine-tune host-pathogen interactions in the airway based on the structure of PGN encountered. Prior studies have also shown that endogenous PGLYRP1 bolstered NOD2-driven macrophage activation in response to GMTriP-K, but not in response to related PGN fragments such as GMDiP or MDP, highlighting the structural specificity of PGLYRP1-mediated immune recognition24. We expand on these findings to suggest exogenous PGLYRP1 biases NOD1 responses and dampens NOD2 responses. Furthermore, we suggest pathogens like B. pertussis have evolved to release PGN structures, such as TCT, which dampen the ability of PGLYRP1 to signal via TREM-1 (Fig. 4C), suggesting an active area of ongoing host-pathogen evolution.

Here, we propose PGLYRP1 as a PGN-structure specific regulator of inflammatory responses. This is comparable to SIGLECs, which can distinguish self from non-self glycans to ignore or elicit immune responses34. During B. pertussis infection, PGLYRP1, predominantly expressed by neutrophils, contributes to early bacterial clearance, as PGLYRP1 KO mice exhibit significantly higher bacterial loads at 4DPI. The bactericidal activity of PGLYRP1 is diminished by the presence of extracellular polysaccharide, highlighting the potential for Bordetella pertussis-derived extracellular decoys to prevent antibiotic activity. PGLYRP1 KO mice exhibit elevated inflammatory cytokine expression, suggesting PGLYRP1 also serves an anti-inflammatory function. Transcriptomic analysis of PGLYRP1 KO lungs infected with mutants which fail to release TCT or the parental strain support the hypothesis that PGLYRP1 limits inflammation in a PGN structure-dependent manner: specifically, the monomeric PGN fragment TCT potentiates the anti-inflammatory activity of PGLYRP1. Mechanistically, PGLYRP1 enhances NOD1 signaling in response to TCT while suppressing NOD2 signaling in response to MDP directing the polarization of immune responses away from a more inflammatory NOD2-driven response. Single cell RNA sequencing revealed that PGLYRP1 expression is enriched in NOD1+ neutrophils, which have less inflammatory profiles than NOD2+ neutrophils. These data suggest B. pertussis may exploit PGLYRP1 to amplify a NOD1-dominant signaling environment that blunts host inflammation. Recent work from our group highlighted the ability of TCT to polarize NOD responses towards NOD1 as an immune evasion mechanism18. This work expands on this finding, highlighting the role of PGLYRP1 and PGN structure in these findings

These findings shift our understanding of mammalian PGLYRP1 from a passive bactericidal protein to an immunomodulator, whose contribution to inflammation is determined by the structure of PGN it encounters. Additionally, they inform our views of PGN structure and how it impacts host responses, suggesting PGN release, particularly 1,6-anhydroMurNAc containing PGNs, as a mechanism to dampen host immunity. These results propose a novel paradigm in host-pathogen interactions; that bacteria may release structurally tailored PGN fragments to bias host immune sensors and signaling in their favor.

Excitingly, we report a novel axis by which PGLYRP1 regulates TREM-1, a known amplifier of inflammation. While PGLYRP1 synergized with PGNs from S. aureus and a B. subtilis to enhance TREM-1 activation, the presence of TCT antagonized this effect (Fig. 4). There are subtle differences in the structures of these PGNs: B. subtilis PGN and TCT both have a DAP-type peptide stem, but the former contains an additional terminal Alanine (pentapeptide) and lacks the 1,6-anhydro bond ring on the MurNAc sugar, which is present in TCT. S. aureus contains a Lys-type peptide stem and supported PGLYRP1 mediated-activation of TREM-1. This suggests that it could be the glycan region of TCT which confers the observed differences in PGLYRP1-PGN mediated activation of TREM-1. Bifidobacterium have been predicted to produce abundant anhydro-PGNs which markedly reduce LPS-stimulated cytokine expression in RAW264.7 cells, underscoring not only the anti-inflammatory effects of 1,6-anhydro-MurNAc containing PGNs but also their broader immunomodulatory potential35.

Previous work from our group demonstrated that inhibition of TREM-1 prevented immunopathology and inhibited cytokine and chemokine responses to infection, highlighting the potential utility of disturbed TREM-1 signaling for B. pertussis33. TREM-1 inhibitor treated mice showed reduced inflammation and inflammatory cytokine responses following B. pertussis infection highlighting the importance of this pathway in pertussis pathogenesis and the potential utility of its manipulation by bacteria. Further, our findings suggest that selectively targeting TREM-1 or PGN structure sensing by PGLYRP1 could attenuate disease severity without compromising bacterial clearance, a strategy aligned with emerging host-directed therapeutics in TB and sepsis36-38.

We hypothesize that B. pertussis may attenuate host immune responses not merely by evading recognition, but by actively hijacking host regulatory circuits through the release of specific PGN structures. Taken together, our findings support a model in which PGLYRP1 serves dual and temporally distinct roles during infection, by promoting bacterial killing while restraining inflammation in a PGN-structure specific manner. This duality may reflect temporal or disease stage specific changes in the types of PGN available to host PRRs, as 1,6-anhydro MurNAc containing PGNs are usually internal, unless released following bacterial lysis. The ability of B. pertussis to exploit this shift by releasing TCT to suppress excessive inflammation represents a novel immune evasion strategy.

More broadly, this work demonstrates that the structure of PGN fragments is a key determinant of host immune outcomes. Innate receptors like PGLYRP1 relay this structural information to shape both defense and pathology, providing an avenue for pathogen hijacking by alternating the availability of different PGN structures. Targeting the TCT–PGLYRP1–TREM-1 axis may offer a novel therapeutic or vaccine-mediated approach to mitigate immunopathology in pertussis and potentially other infections involving secreted PGN. These findings reframe PGN as both a pathogen-associated molecular pattern and an immunomodulatory cue that both host and pathogen exploit to regulate the immune-inflammatory environment. While our data implicate the 1,6-anhydro MurNAc moiety in TCT as key to its immunomodulatory activity, future studies using defined synthetic muropeptides will be necessary to pinpoint structural determinants. Likewise, how these alterations in PGN structure impact PGLYRP1 interactions with host PRRs remains unclear.

Methods and materials

Mice and Ethics Statement

All animal procedures were approved by the Institutional Animal Care and Use Committee at the University of Maryland School of Medicine (protocol 00000108). Adult (6–8-week-old) BALB/c and C57BL/6, PGLYRP1 knockout (PGLYRP1 KO), and TREM-1 knockout (TREM-1 KO) mice were bred and housed under specific pathogen-free conditions. Neonatal pups (5–7 days old) were infected by aerosol (OD1.0, 20 minutes) and monitored daily for survival, weight, and clinical signs of disease for up to 14 days. Humane endpoints were applied in accordance with IACUC guidelines.

Bacterial Strains and Infections

B. pertussis strains, including wild-type BC36, a streptomycin-resistant Tohama I derivative, or isogenic mutants (TCT-, TCT+, BpsB KO) were cultured on Bordet-Gengou agar supplemented with 10% defibrinated sheep blood or in Stainer-Scholte broth. TCT mutants were generously supplied by William Goldman (University of North Carolina at Chapel Hill). TCT-strain was generated by expression of the E. coli AmpG in BC36, resulting in a 50-fold decrease in TCT release. TCT+ was generated by replacing native B. pertussis AmpG with a kanamycin resistance cassette, this resulted in a 24-fold increase in TCT release30. Adult mice were anesthetized and inoculated intranasally with 2 × 106 CFU of B. pertussis in 50 μL PBS. Lungs were harvested at 4DPI and 7DPI for bacterial burden analysis by plating of serially-diluted lung homogenates.

RNA isolation and processing

Lung tissue was harvested at 2 or 7DPI for RNA isolation by the Trizol-choloroform method. Briefly, lung tissue was homogenized using the Omni Bead Beader (Omni, Inc), phase separated using chloroform, and precipitated using isopropanol. RNA was then quantified and converted to cDNA (iScript cDNA Synthesis Kit). Quantitative real-time PCR (qPCR) was performed using the BIORAD CFX96 real-time PCR instrument. Gene expression was calculated as fold change relative to PBS-inoculated control animals using the 2^−ΔΔCT threshold cycle method and normalized to Hypoxanthine phosphoribosyltransferase (HPRT) (internal housekeeping gene).

Bulk RNA Sequencing

At 4DPI, lungs were harvested from 3 WT or PGLYRP1 KO BALB/c mice infected with WT or TCT-deficient B. pertussis. Total RNA was extracted from homogenized lungs using Trizol Reagent (Invitrogen) according to the manufacturer’s protocol. RNA integrity was confirmed using a Bioanalyzer (Agilent), alignments were generated by HISAT2 sequencing performed using a Illumina NextSeq 1000. FastQC and RSeQC were used to assess read and alignment quality. Samtools was used to generate alignment statistics. Reads were aligned with the Mus musculus GRCm39 reference genome. Differential gene expression was analyzed using DESeq2, with significance defined as adjusted p-value < 0.05.

Single-Cell RNA Sequencing

Single-cell suspensions were prepared from lung tissues of duplicate WT and PGLYRP1 KO C57BL/6 mice at 4DPI. Following red blood cell lysis, cell viability was assessed by trypan blue staining, viable cells were sorted and loaded onto the 10x Genomics Chromium Controller for barcoding and library construction. Sequencing was performed on an Illumina NovaSeq platform. Data were processed using Cell Ranger and analyzed using the Seurat package in R. Quality control cutoffs of between 1000 and 10000 features and <20% mitochondrial DNA. Data were normalized and scaled in Suerat v5. Clusters were identified, determined, and visualized via Seurat library workflow (RunPCA, FindNeighbors, FindClusters, RunUMAP) and then annotated via hallmark markers determined by differential gene expression analysis across clusters using Wilcoxon rank-sum tests.

Neutrophil Isolation and Bactericidal Assays

Bone marrow-derived neutrophils (BMDNs) were isolated by density gradient centrifugation from adult C57BL6 mice and cultured in DMEM supplemented with 10% FBS. For killing assays, BMDNs were incubated with B. pertussis at a multiplicity of infection (MOI) of 10:1 for 2 or 24 hours. Supernatants were serially diluted and plated to quantify surviving CFU. Recombinant murine PGLYRP1 (R&D Systems) or BSA control was incubated with live B. pertussis or purified PGN fragments (TCT gifted by Prof. William Goldman or S. aureus PGN, Invivogen) to assess direct bactericidal activity.

RNAscope In Situ Hybridization

RNA in situ hybridization was performed on 5um formalin fixed paraffin embedded lung sections with the RNAscope detection kit to confirm RNA sequencing identification of PGLYRP1 transcripts (Advanced Cell Diagnostics). Hybridization, amplification, and detection steps were carried out according to the manufacturer’s protocol. Briefly, slides were deparaffinized in xylene and rehydrated using an ethanol series. Sections were treated with hydrogen peroxide for 10 minutes at room temperature, followed by treatment with target retrieval reagent (Advanced Cell Diagnostics). Sections were hybridized with target-specific probes (Advanced Cell Diagnostics) followed by amplification and fluorescent detection (Opal 570 and Opal 520). Slides were counterstained with hematoxylin and eosin and imaged using an ECHO Revolve 4 and analyzed with ImageJ.

Reporter Cell Assays for NOD1, NOD2, and TREM-1 Activation

HEK293-derived reporter cells expressing murine or human NOD1 and NOD2 with an, or TREM-1 and an NFkB–dependent luciferase or NF-kB inducible SEAP reporter (InvivoGen) and TREM-1-DAP12 reporter cells (Eurofins Discover X)in the Jurkat T cell background with an NFAT-dependent luciferase were were stimulated with B. pertussis tracheal cytotoxin TCT (generously provided by William Goldman, UNC), PGN fragments (TCT, S. aureus or B. subtilis PGN (Invivogen)) in the presence or absence of recombinant human PGLYRP1 (R&D systems) at a concentration of 5ug per well for 24 hours. Supernatants were analyzed for reporter activity using colorimetric (NOD1 and NOD2 reporter cells; Promega) or luminescent (TREM-1 reporter cells; Invivogen) detection kits (Promega).

Histopathology

Lungs were fixed in 10% formalin, embedded in paraffin, and sectioned at 5 μm thickness. Slides were stained with hematoxylin and eosin (H&E). A blinded veterinary pathologist scored sections for peribronchial and perivascular inflammation, alveolitis, and airway epithelial damage.

eLife Assessment

Rickert and colleagues demonstrate that the host peptidoglycan-binding protein PGLYRP1 has both beneficial and detrimental effects on Bordetella pertussis infection in mice. Using a solid array of techniques, the study provides useful insights into how peptidoglycan species may alter host immune responses. The data on the bactericidal effects on B. pertussis are incomplete, and further experiments are needed to draw conclusions on this question.

Reviewer #1 (Public review):

Summary:

The authors aim to demonstrate that PGLYRP1 plays a dual role in host responses to B. pertussis infection. PGLYRP1 signaling is known to activate bactericidal responses due to recognition of peptidoglycan. Through NOD1 activation and TREM-1 engagement, it appears PGLYRP1 also has immunomodulator activities. The authors present mouse knockout studies and gene expression data to illustrate the role of PGLYRP1 in relation to B. pertussis peptidoglycan. Mice lacking PGLYRP1 had slightly lower pathology scores. When TCT peptidoglycan was removed from the bacteria, surprisingly IL23A, IL6, IL1B, and other pro-inflammatory genes encoding cytokines increased. The relationship to TCT and PGLYRP1 suggests the pathogen uses this strategy to decrease immune activation. The authors went on to show the relationship between PGLRP1 and TREM-1 as mediated by PGN using various versions of peptidoglycan. The study presents multiple angles of data to back up its findings and demonstrates an interesting strategy used by B. pertussis to downregulate innate responses to its presence during infection.

Strengths:

Use of knockout mice of the key factor being considered, paired with isogenic B. pertussis strains, to reveal the mechanism of immune modulation to benefit the bacteria. The authors used in vivo gene expression paired with in vivo assays to establish each aspect of the mechanism.

Weaknesses:

The main focus was on innate responses, and some analysis of antigen-specific antibody responses could improve the impact of the findings.

Reviewer #2 (Public review):

Since its original discovery, the mechanistic basis for TCT-mediated pathogenesis of Bordetella pertussis has been a moving target and difficult to uncouple from confounding variables. The current study provides some exciting data that suggest PGLYRP-1 modulates host responses upon 'activation' by TCT. While there are some strengths associated with the unbiased approaches and collective data to support the claims associated with TCT and PGLYRP-1's function in this system, caution should be used when interpreting and extrapolating some of the information provided. For instance, the amount and purity of TCT used in the studies are unclear, and the in vitro activity of PGLYRP1 on B. pertussis is questionable. Different mouse backgrounds are used for various assays throughout, and it is known that the PRRs vary in these systems, so the confounding variables are difficult to uncouple. Additional concerns include the types of statistical tests being performed to support some of the claims and the relevance of using whole, intact PG sacculi from other species for comparative studies with a fragment of released PG (i.e., TCT).

Reviewer #3 (Public review):

Summary:

This study evaluates the contributions of the mammalian PG-binding protein PGLYRP1 to Bordetella infection. The authors find potential roles for PGLYRP1 in both bacterial killing (canonical) and regulation of inflammation (non-canonical). While these are interesting findings and the idea that PG fragment release has differential impacts on infection depending on fragment structure, the study is limited by the lack of connection between the in vivo and in vitro experiments, and determining the precise mechanism of how PGLYRP1 regulates host responses and bacterial fitness during infection requires further study.

Strengths:

(1) The combination of scRNAseq with in vitro and in vivo assays provides complementary views of PGLYRP1 function during infection.

(2) The use of TCT-deficient B. pertussis provides a useful control and perturbation in the in vitro assays.

Weaknesses:

(1) The study does not ultimately resolve the initial early versus late phenotype divergence. While the in vitro assays suggest explanations for their in vivo observations, further mechanistic links are lacking and necessary for the author's conclusions throughout. To state one example, what is the early and late infection phenotype of TCT- Bp in mice lacking PGLYRP1? RNAseq data are reported from these mice, but there are no burden or pathology studies. Furthermore, what are the neutrophil phenotypes (NOD-1/TREM-1 activation) in vivo? And are they dependent on PGLYRP1 and/or TCT?

(2) It is unclear whether or how the NOD1 and TREM-1 pathways interact.

(3) Many of the study's conclusions rely on the use of HEK293 reporter lines in the absence of bacterial infection, which may not be physiologically representative.

(4) The methods lack detail overall, and the experimental procedures should be described more concretely, especially for the scRNAseq datasets.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

The authors aim to demonstrate that PGLYRP1 plays a dual role in host responses to B. pertussis infection. PGLYRP1 signaling is known to activate bactericidal responses due to recognition of peptidoglycan. Through NOD1 activation and TREM-1 engagement, it appears PGLYRP1 also has immunomodulator activities. The authors present mouse knockout studies and gene expression data to illustrate the role of PGLYRP1 in relation to B. pertussis peptidoglycan. Mice lacking PGLYRP1 had slightly lower pathology scores. When TCT peptidoglycan was removed from the bacteria, surprisingly IL23A, IL6, IL1B, and other pro-inflammatory genes encoding cytokines increased. The relationship to TCT and PGLYRP1 suggests the pathogen uses this strategy to decrease immune activation. The authors went on to show the relationship between PGLRP1 and TREM-1 as mediated by PGN using various versions of peptidoglycan. The study presents multiple angles of data to back up its findings and demonstrates an interesting strategy used by B. pertussis to downregulate innate responses to its presence during infection.

Strengths:

Use of knockout mice of the key factor being considered, paired with isogenic B.

pertussis strains, to reveal the mechanism of immune modulation to benefit the bacteria. The authors used in vivo gene expression paired with in vivo assays to establish each aspect of the mechanism.

Weaknesses:

The main focus was on innate responses, and some analysis of antigen-specific antibody responses could improve the impact of the findings.

The authors thank the reviewer for their careful reading of the manuscript. We agree that understanding the impact of peptidoglycan recognition in adaptive immunity, including antibody responses, would be beneficial. This is particularly apparent due to the pressing need for novel vaccination strategies for pertussis. To this end, we have modified the discussion section to highlight this and are embarking on detailed studies of the adaptive response generated with B. pertussis strains releasing alternative peptidoglycan structures.

Reviewer #1 (Recommendations for the authors):

(1) This reviewer is of the opinion that describing the PGLYRP1 as a "bactericidal protein" seems misleading. "To determine whether PGLYRP1 has bactericidal activity against B. pertussis, we performed in vitro and ex vivo killing assays." Bactericidal activity was measured in normal or knockout neutrophils, but this seems to say the PGLYRP1 itself is an antimicrobial peptide. It clearly plays a role in the response,e but it is a regulator and not a killing agent.

We agree that ‘bactericidal’ is not the most accurate description and have revised the manuscript accordingly to be more accurate throughout results section 1.

(2) PGN can induce IgM production. Antibody production of any type was absent from this study. Would IgA/IgM/IgG levels to B. pertussis or its TCT change due to PGLYRP1? To this reviewer, it would be good to use the serum and perform some ELISA analysis. It is also likely that T cell responses could be impaired, but that may be out of the scope of this manuscript, but could be acceptable to consider for future studies.

The authors thank the reviewers for this suggestion. We have added text to the discussion section to highlight the importance and potential of this suggestion.

(3) Please include sources of mice (vendors) and strain numbers for transparency.

The authors have added the relevant detail to the methods section to address this valid concern.

(4) Were female or male mice or both used?

For PGLYRP1 vs BALB/c comparisons both male and female mice were used. These are presented as combined data. No discernible differences were noted between male and female mice following infection. For single cell RNA sequencing studies only female mice were used, to be consistent with the published pertussis mouse model and avoid sex-based complications in analysis. We have clarified these details in the text.

(5) It appears B. pertussis was cultured in SSM or BG. What condition was used for the bacteria used for the mouse challenge? SSM or BG?

For mouse studies, bacteria were grown on BG agar supplemented with 10% defibrinated sheep blood for 48 hours and inoculum prepared by suspending in PBS in accordance with our established protocols. For in vitro studies liquid cultures were grown to mid-log in SSM. This has now been clarified in the methods section

(6) Are the raw RNAseq and scRNAseq reads deposited in SRA?

Raw data has now been deposited in the Gene Expression Omnibus (GEO) under number GSE324217

(7) Is the scRNAseq data from one mouse or a pooled set of mice? If pooled, were the individual mice barcoded?

scRNAseq data was obtained from barcoded individual samples and replicates were pooled and integrated during analysis, but the individual mouse each cell came from is still noted in the downstream analysis. This is now clarified in methods.

(8) Why were some studies done by aerosol and others were done by intranasal delivery?

The authors thank the reviewer for careful reading of the manuscript which erroneously listed aerosol infections. All infections in these studies were intranasal. This has now been rectified in the text.

Reviewer #2 (Public review):

Since its original discovery, the mechanistic basis for TCT-mediated pathogenesis of Bordetella pertussis has been a moving target and difficult to uncouple from confounding variables. The current study provides some exciting data that suggest PGLYRP-1 modulates host responses upon 'activation' by TCT. While there are some strengths associated with the unbiased approaches and collective data to support the claims associated with TCT and PGLYRP-1's function in this system, caution should be used when interpreting and extrapolating some of the information provided. For instance, the amount and purity of TCT used in the studies are unclear, and the in vitro activity of PGLYRP1 on B. pertussis is questionable. Different mouse backgrounds are used for various assays throughout, and it is known that the PRRs vary in these systems, so the confounding variables are difficult to uncouple. Additional concerns include the types of statistical tests being performed to support some of the claims and the relevance of using whole, intact PG sacculi from other species for comparative studies with a fragment of released PG (i.e., TCT).

We thank the reviewer for their insightful suggestions to improve the standard of our manuscript and for highlighting several important considerations regarding our interpretation of TCT mediated host responses. We have addressed the points made in the revised manuscript. In particular, we have amended the Methods section to include a description of the purification and quantification of tracheal cytotoxin. These additions clarify the dosing of TCT used throughout the manuscript. We have revised the Results and Discussion sections to avoid overstating the bactericidal activity of PGLYRP1 against B. pertussis and to more carefully describe in vitro observations. Our revised interpretation emphasizes the role of PGLYRP1 in modulating host immune responses. Additionally, we have clarified experimental design and strain usage descriptions in the Methods section. The reviewer provided valuable and insightful comments on the solubility and structure of muropeptides studies. In response, we have revised the Results and Discussion sections to acknowledge these differences and the limitations they pose. Further, we have removed conclusions regarding the specific role of the 1,6anhydro bond. The statistical analyses have been reviewed and validated as well as clarified throughout the manuscript and Methods and figure legends updated.

We appreciate the reviewer’s comments and believe the revisions have improved the clarity and rigor of the manuscript while maintaining the central conclusions about how peptidoglycan recognition influences host inflammatory responses during B. pertussis infection.

Reviewer #2 (Recommendations for the authors):

Major Points:

(1) The concentration, purity, etc. of TCT seems like it is entirely unknown. Only a couple of experiments actually state the amount used, and it's unclear how the author determined the concentration because this is not trivial. Given the long-standing concerns with purity and co-purifying contaminants, this issue is paramount and needs to be properly addressed.

TCT was purified by HPLC in the Goldman lab (UNC). Concentration was determined by comparing the peak area of each preparation to a purified TCT standard quantified by amino acid analysis. We have added these details to the Methods and now report concentrations throughout the manuscript.

(2) Related to the effects of bacterial PG, studies performed are comparing TCT (a muropeptide) to commercially acquired, insoluble PG sacculi from B. subtilis and S. aureus. One cannot make these comparisons. There are flaws in terms of solubility (one goes into solution, the other does not), the amount used, the molar concentrations, etc. The authors also state that these are non-1,6 anhydro PG samples. That is not true. They contain plenty of 1,6 anhydroMurNAc, the moiety just exists in a different form. Finally, B. subtilis PG is not just mDAP, it's amidated, which is known to have effects on host response(s).

We thank the reviewer for this important critique. We agree that differences in solubility and structural composition between TCT and PG sacculi limit direct comparisons. We have revised the Results and Discussion to remove statements implying direct equivalence and instead frame these experiments as highlighting how structural and physical properties of PGN fragments influence PGLYRP1-mediated activation of TREM-1. We have also removed statements regarding the 1,6-anhydro bond which were not adequately supported.

(3) The claim that PGLYRP-1 is bactericidal in vitro is not supported by the data. Figure 1G shows that 24 hours after incubation, there is no difference. The comparison is being made to BSA, which is much higher (possibly because they're catabolizing it?) and thus entirely inappropriate. All other data in Figure 1 suggest no effect in vitro. In fact, it's this reviewer's position that none of the studies in Figures 1G, H, and I are convincing and should be entirely excluded.

The authors agree that language describing the bactericidal assays is not optimal and have made revisions. The text in this results section has been modified to more carefully describe bacterial killing assays and accurately describe the effects the data suggest, primarily removing claims of bactericidal effects. BSA was chosen as a control protein (concentration matched with PGLYRP1), based on published controls for PGLYRP bactericidal assays (Lu et al 2006, JBC) similar results were obtained with PBS (volume matched with PGLYRP1). Descriptions of Fig1G,H,I have been updated. Data in 1H demonstrates that TCT release does not protect against effects of PGLYRP1, despite free PGN inhibiting PGLYRP1 bactericidal activity in published literature, while 1I suggests that extracellular polysaccharides contribute to protection against PGLYRP1 activity, preventing a more bactericidal phenotype which were not observed in the earlier assays when B. pertussis retained its capacity to produce bps polysaccharide.

(4) Histology studies are unclear, and the data presented do not support the claims. Not only are the methods and results text describing the analysis contradictory, but nowhere are the actual statistical tests supporting the claims that they are different provided. This might be an oversight, but based on the variation, I would be surprised if they were statistically significantly different if proper tests are being used.

Significance for pathology scores were initially determined using 2-way ANOVA as we had 4 groups (WT&KO at 4&7DPI) providing p-values of 0.01 for WT vs KO at 7DPI and 0.003 at 4DPI. Following reviewers’ suggestions, we have reanalyzed these data using a Mann Whitney U test, which is more appropriate for comparisons between two groups. This analysis yielded p-values of 0.013 (4DPI) and 0.00316 (7DPI) respectively confirming that the observed differences remain statistically significant. Statistical methods are now described in the methods and figure legends.

(5) The NOD reporter studies are not well controlled and should include a) mouse vs human for both NOD1 and NOD2; b) defined details in terms of how spent culture media was treated, amount of material normalized, etc., c) concentrations of all materials used.

We appreciate the reviewer’s comments regarding the NOD reporter assays. In response: (a) We have clarified and articulated the murine/human NOD reporter assays and included both human and mouse NOD1, along with controls. (b) We have supplemented descriptions of how conditioned (spent) culture media were collected, processed, and normalized in the ‘Bacterial strains and infections’ and ‘Reporter Cell Assays’ methods sections; (c) and the final concentrations of all agonists and test materials used in the reporter assays are now specified in the Methods and corresponding figure legends. Together, these additions address the requested controls and clarify the experimental conditions

(6) The scRNA-seq studies are provocative and informative, but the data shown are selectively included for the purposes of the paper. This is justified in terms of 'telling a story', but it's a disservice to the community not to include all the raw data attained. These should be deposited in an open-source system.

The complete dataset has now been deposited in GEO (GSE324217) enabling full access for the community. The analyses presented in the manuscript focus on the datasets most relevant to the central conclusions.

Minor points:

(1) The authors refer to arthropod PGRPs but call them PGLYRPs. It is best to stick with the established nomenclature and use the proper names to distinguish each. There are a few sentences in the abstract that don't make sense as they're written.

The authors thank the reviewer for their careful reading of the manuscript and have altered the manuscript to use PGRP for arthropod peptidoglycan recognition proteins.

(2) The reciprocal result of bacterial burden at different time points in the context of PGLYRP-1 production in mice could be simply explained - it is bactericidal early, and the accumulation of dead/dying bacteria releases large pieces of PG that are not released during growth (anhydro) but rather lysis. It is the latter that causes the inverse relationship later.

The authors believe this is an interesting and plausible explanation for differences in responses at different stages of disease. Further, we believe that elucidating the mechanism by which ‘large pieces of PG not released during growth” are recognized differently than PG from lysed bacteria is worthwhile. We speculate that the release of TCT could be a mechanism by which B. pertussis takes advantage of host differences in PG recognition. We thank the reviewers for this thought and have included this possible interpretation in the text.

(3) The results section references Figure 1G while discussing results presented in Figure 1H.

This has now been corrected.

Reviewer #3 (Public review):

Summary:

This study evaluates the contributions of the mammalian PG-binding protein PGLYRP1 to Bordetella infection. The authors find potential roles for PGLYRP1 in both bacterial killing (canonical) and regulation of inflammation (non-canonical). While these are interesting findings and the idea that PG fragment release has differential impacts on infection depending on fragment structure, the study is limited by the lack of connection between the in vivo and in vitro experiments, and determining the precise mechanism of how PGLYRP1 regulates host responses and bacterial fitness during infection requires further study.

Strengths:

(1) The combination of scRNAseq with in vitro and in vivo assays provides complementary views of PGLYRP1 function during infection.

(2) The use of TCT-deficient B. pertussis provides a useful control and perturbation in the in vitro assays.

Weaknesses:

(1) The study does not ultimately resolve the initial early versus late phenotype divergence. While the in vitro assays suggest explanations for their in vivo observations, further mechanistic links are lacking and necessary for the author's conclusions throughout. To state one example, what is the early and late infection phenotype of TCT- Bp in mice lacking PGLYRP1? RNAseq data are reported from these mice, but there are no burden or pathology studies. Furthermore, what are the neutrophil phenotypes (NOD-1/TREM-1 activation) in vivo? And are they dependent on PGLYRP1 and/or TCT?

(2) It is unclear whether or how the NOD1 and TREM-1 pathways interact.

(3) Many of the study's conclusions rely on the use of HEK293 reporter lines in the absence of bacterial infection, which may not be physiologically representative.

(4) The methods lack detail overall, and the experimental procedures should be described more concretely, especially for the scRNAseq datasets.

We thank the reviewer for their comprehensive and fair assessment of our study and for highlighting both its strengths and areas where clarification could improve the manuscript. As noted in the review the possibility that peptidoglycan fragment structure impacts disease pathogenesis is interesting and the role of PGLYRP1 in regulating host and bacterial fitness during infection requires further study.

We have addressed the points made by the reviewer in the revised manuscript. We edited the Methods section to provide additional experimental detail, particularly for the scRNA-seq analyses and reporter assays. We also clarified the experimental design and interpretation of the in vitro studies to avoid overstating mechanistic conclusions.

Studies with TREM-1 and NOD are attempting to assess multiple aspects of PGN/PGLYRP mediated enhancement of inflammatory responses via NFkB/MAPKs. No attempts have been made to assess synergistic, overlapping or compensatory effects between these systems. Other work from our group highlights the role of peptidoglycan in driving inflammatory responses via NOD receptors (doi: https://doi.org/10.1101/2025.08.08.669383) and TREM-1 (doi: 10.1128/IAI.00126-21). Work in this paper assesses the contribution of these pathways to the observed immune modulation noted by PGLYRP1.

We have clarified figure legends and analyses, including interpretation of neutrophil transcriptional programs identified in scRNAseq datasets and comparisons to known neutrophil phenotypes.

We appreciate the reviewers feedback and the opportunity to improve the clarity of our manuscript and optimize the conclusions and central findings.

Reviewer #3 (Recommendations for the authors):

(1) Please clarify in Figure 1C what the axis means, since the text refers to both uninfected and infected cells. What data allow the conclusion that PGLYRP1 expression "expanded" to other cell subsets?

We thank the reviewers for catching this oversight. We were relying on data which we had not best represented in Figure 1C, so we updated this figure and corresponding text so that this violin plot demonstrates increased PGLYRP1 expression levels and an increasing or expanding number of cell types following infection. This is now also reflected in the text. Expression of PGLYRP1 is apparent in more cell types and to a greater extent following infection (red) with B. pertussis compared to PBS challenge (black). Expression represents normalized and transformed unique molecular identifier counts per gene per cell.

(2) Please revise the Figure 1 legend to match the Figure panels, and mention the time point of the mPGLYRP1 killing assay in 1H/I. Were these assays performed at 6 or 24 hours? This could affect the interpretation of the data.

This has been revised to reflect timing of data.

(3) The text at the end of the first Results section is overstated, as the data in Figure 1 do not relate to immune-mediated clearance apart from expression levels.

This text has been revised and reference to immune mediated clearance removed

(4) More detail is needed in the explanation of Figures 3E-G. Do the neutrophil subsets correspond to known subsets from the literature?

When we overlaid established neutrophil signatures from the literature onto our dataset the NOD2+ neutrophils most closely resembled inflammatory or activated neutrophil programs described previously (Xie et al. 2020 Nat. Immuno., Veglia et al. 2021 J. Exp. Med)- specifically, high il1a, Ccl3 and Ptgs2 expression. In contrast, NOD1+ neutrophils showed greater overlap with resolving or regulatory neutrophil states- including genes associated with lipid mediator metabolism and NFkB dampening. Importantly, the clustering itself was not driven by NOD1 or NOD2 expression alone. NOD expression segregated within transcriptionally distinct neutrophil programs that are consistent with previously described inflammatory versus regulatory subsets. We included descriptions of these inflammatory neutrophils and related them to previously identified neutrophil populations, supporting our findings and improving the representation and articulation of the single cell neutrophil data analysis. We deeply thank the reviewers for their help in improving this section.

(5) The Methods section describes qPCR, but this is not presented in the Results.

This has now been removed. We thank the reviewer for their careful and complete review of the manuscript.

Version 2

Abstract

Bordetella pertussis, the etiologic agent of whooping cough, remains a serious public health concern despite widespread vaccination. Improved therapeutics and vaccines are urgently needed to treat and prevent pertussis disease. Host recognition of bacterial peptidoglycan (PGN), including B. pertussis extracellular PGN fragment tracheal cytotoxin (TCT), shapes the immune response to infection. Peptidoglycan recognition proteins (PGLYRPs) are a conserved family of innate immune molecules which bind bacterial PGN. While they function as immune signaling receptors in arthropods (termed PGRPs in arthropods), PGLYRPs in mammals have thus far been primarily recognized for their bactericidal activity. Previously thought to function only as antimicrobial peptides in mammals, the immune modulatory roles of this family of peptidoglycan recognition proteins are beginning to gain greater appreciation. Peptidoglycan recognition protein 1 (PGLYRP1) is a secreted antimicrobial protein. However, its role in mammalian host defenses and immune signaling during infection with Gram-negative pathogens, such as B. pertussis, remain largely unknown. Here, we identify a dual role for PGLYRP1 in modulating host immune responses to B. pertussis. Using knockout mice, single-cell and bulk transcriptomics and functional assays, we show that PGLYRP1 contributes to host antibacterial responses to B. pertussis. PGLYRP1 also dampens inflammatory responses and paradoxically inhibits bacterial killing later in infection. Mechanistically, PGLYRP1 enhances nucleotide oligomerization domain (NOD)-1 signaling in response to TCT while suppressing NOD2- and triggering receptor expressed on myeloid cells-1 (TREM-1)-mediated inflammatory pathways. TCT-bound PGLYRP1 selectively impairs TREM-1 activation compared to PGNs from other bacteria, revealing a novel bacterial immune evasion strategy. These findings demonstrate that B. pertussis co-opts PGLYRP1 to temper inflammation and alter immune signaling, revealing a novel immune evasion mechanism of manipulating the availability and structure of their exogenous peptidoglycan, revealing implications for host-pathogen evolution, vaccine design and host-directed therapeutics.

Introduction

Bordetella pertussis is a highly contagious respiratory pathogen and the causative agent of whooping cough, a disease characterized by severe, prolonged coughing and airway inflammation1-3. Despite widespread vaccination efforts, B. pertussis remains a public health concern due to waning vaccine-induced immunity and the pathogen’s capacity to evade host immune defenses4. Host detection of microbial components, including peptidoglycan fragments, is essential for initiating immune responses to infection. Understanding how these host-pathogen interactions shape disease outcomes is critical for developing improved strategies to fight infections.

Peptidoglycan (PGN) is a fundamental component of the bacterial cell wall, composed of repeating disaccharide chains cross-linked by short peptides5. PGN structure is an important determinant of innate immune recognition and responses6, 7. During cell wall turnover, some pathogenic bacterial species, including B. pertussis, release PGN fragments into the extracellular environment8-10. The biological roles of these fragments in infection and immunity are poorly defined.

B. pertussis releases a disaccharide-tetrapeptide monomeric PGN fragment known as tracheal cytotoxin (TCT)11, 12. TCT differs from B. pertussis cell wall PGN by having a 1,6-anhydro group on its MurNAc sugar due to incomplete PGN turnover11. The functional consequences of this 1,6-anhydro MurNAc modification are poorly understood13. TCT is recognized by the cytosolic pattern recognition receptor (PRR) NOD1 to activate NFkB-driven immune responses14, 15. In vitro and ex vivo studies suggest that recognition of TCT by NOD1 in non-ciliated mucus-secreting cells leads to the accumulation of nitric oxide within neighboring ciliated airway epithelial cells and subsequent ciliostasis and extrusion of ciliated cells16, 17. We recently demonstrated that TCT has broader immunomodulatory effects, skewing PGN-sensing toward NOD1 and away from NOD2, thereby dampening a stronger NOD2-dependent, pro-inflammatory cytokine production and generation of immune memory18. However, the broader immunological consequences of TCT release during infection, especially in the context of PGN-sensing by PRRs remain poorly defined.

Peptidoglycan Recognition Proteins (PGLYRPs) are a conserved family of soluble extracellular innate immune receptors that bind PGN19. In arthropods, PGRPs act as extracellular pattern recognition receptors (PRR) that trigger downstream immune pathways20. In mammals, they have been characterized as bactericidal proteins that target PGN on Gram-positive bacterial surfaces21, 22. However, recent studies have expanded their role beyond antimicrobial defense, suggesting function in regulating host immune responses. PGLYRP1 is a secreted protein stored in neutrophil granules, that is directly bactericidal against several bacterial species23. Recent work suggests that mammalian PGLYRPs also regulate adaptive immunity, including CD8+ T cell responses, in autoimmunity and cancer, and act as an accessory protein for the activation of inflammation amplifying receptor TREM-1 and NOD2-dependent signaling24-26.

TREM-1 is a PRR which amplifies inflammatory responses during infection27. Activation of TREM-1 by its ligands, including PGLYRP1-PGN complexes, enhances production of proinflammatory cytokines and chemokines28. Excessive TREM-1 activation is associated with inflammatory pathology in sepsis and other diseases29. Despite growing interest in PGLYRPs, their roles in host-pathogen interactions, particularly in the context of Gram-negative bacteria and secreted PGN fragments, remains poorly defined.

Here, we demonstrate that PGLYRP1 plays a dual role in host responses to B. pertussis infection, contributing to early bactericidal responses and fine-tuning inflammation in response to distinct PGN structures. We demonstrate that exogenous PGLYRP1 enhances NOD1 activation by TCT but failed to enhance muramyl dipeptide (MDP) activation of NOD2. Further, PGLYRP1 directly activates the pro-inflammatory receptor TREM-1, a response that is potentiated by sacculus derived PGN from Gram-positive bacteria Staphylococcus aureus or Bacillus subtilis. PGLYRP1-mediated activation of TREM-1 was not supported by B. pertussis extracellular PGN monomer TCT. These findings reveal that PGLYRP1 is not only contributes to anti-bacterial responses but suggests it may also be a context-specific immunomodulator whose function is shaped by the PGN it encounters. This work uncovers a previously unrecognized mechanism by which extracellular PGN and innate receptors together shape host inflammation during infection in a structure-specific manner.

Results

Opposing roles of PGLYRP1 in early and late host defenses to B. pertussis

PGLYRP1 has known antibacterial activity in vitro, particularly against Gram-positive bacteria23. To investigate whether PGLYRP1 contributes to host microbial defenses against B. pertussis, we intranasally infected adult BALB/c wild-type (WT) and PGLYRP1-knockout (PGLYRP1 KO) mice and assessed lung bacterial burden at 4- and 7-days post infection (DPI). Surprisingly, PGLYRP1 had opposing impacts on bacterial burden at these time points (Fig. 1A&B). At 4DPI, PGYLRP1 KO mice exhibited significantly higher bacterial burdens compared to WT controls (p < 0.01), indicating that PGLYRP1 contributes to early bacterial control. However, by 7DPI PGLYRP1 KO mice had significantly lower bacterial burden compared to WT mice (p < 0.001), suggesting that PGLYRP1 impairs bacterial clearance later in infection, potentially by regulating or modulating host immune responses.

PGLYRP1 has a dual role in host defense against Bordetella pertussis.

Lung bacterial burdens in wild-type BALB/c (WT, black) and PGLYRP1-knockout (PGLYRP1 KO, red) at 4-(A) and 7-(B) days post-infection. Violin-plot visualizing single-cell expression levels of Pglyrp1 across cell types in the lungs of C57/B6 at 4DPI for both infected and sham-challenged mice, assessed via analysis of single-cell RNA sequencing data (C). RNAscope in situ hybridization visualization of Pglyrp1 transcripts in FFPE sections of B. pertussis infected mouse lung tissue (D). Each dot represents a single Pglyrp1 mRNA molecule. Images are presented at 20X and 10X magnification of ex vivo bone marrow derived neutrophils (E). Neutrophils were isolated from BALB/c (WT) or PGLYRP1 KO mice and incubated with B. pertussis for 2 (circles) or 24 (square) hours and CFU enumerated by serial dilution on Bordet Gengou agar. In vitro bacterial killing assay assessing recombinant murine PGLYRP1 (mPGLYRP1, red) bactericidal activity at 6- and 24-hours post-incubation compared to BSA control against WT (G) or a TCT non-releasing strain (TCT-) at 6 hours (H). Enumeration of CFU following incubation of mPGLYRP1 with wild-type B. pertussis (parental) or mutant lacking the BpsB gene of the Bordetella polysaccharide (Bps) operon for 6 hours (I). Data represents CFU as a percentage of starting inoculum recovered at indicated timepoint as assessed by serial dilution of culture on BG agar. Data are presented as mean ± SEM; significance determined by unpaired t-test or ANOVA as appropriate. **, p-value < 0.01, *** p-value < 0.005, **** p-value < 0.001

To identify which cell types express PGLYRP1 following infection, we performed single cell RNA sequencing on lungs from uninfected and B. pertussis-infected C57BL/6 mice at 4DPI. In uninfected lungs PGLYRP1 expression was most prominent in neutrophils along with heterogenous expression in some lymphocyte and inflammatory monocyte (iMON) populations (Fig. 1C). Upon infection, the PGLYRP1-expressing cell-types expanded to include epithelial cells like ciliated cell, secretory cells, and AT1/AT2s in addition to granulocytes like basophils (Fig. 1C). RNAscope in situ hybridization confirmed expression in the airways of infected mice and in neutrophils (Fig. 1D, E).

To determine whether PGLYRP1 contributes to antibacterial responses to B. pertussis, we performed in vitro and ex vivo killing assays. Based on transcriptomic data (Fig. 1C), we assessed the bactericidal potential of BMDNs from WT and PGLYRP1 KO mice incubated with B. pertussis for 2 or 24 hours. At 24 hours, PGLYRP1 KO neutrophils failed to control viable bacteria compared to WT neutrophils (Fig. 1F), indicating that PGLYRP1 contributes to neutrophil-mediated control of B. pertussis. In a cell-free assay, recombinant murine PGLYRP1 (mPGLYRP1) significantly reduced bacterial CFU after 24 hours of incubation compared to BSA control (Figure 1G, p<0.001), validating its potential antibacterial role in neutrophil mediated control of B. pertussis growth.

TCT release by B. pertussis is a product of inefficient recycling by the permease AmpG11. Replacing B. pertussis AmpG, with E. coli AmpG resulted in a strain (TCT-) which releases 99% less TCT (TCT(-)), while deletion of AmpG created a strain with 24-fold greater TCT release (TCT(+))30. Since PGLYRP1 binds bacterial PGN and the addition of excess PGN has been shown to prevent PGLYRP bactericidal activity31,,we hypothesized that B. pertussis TCT, an extracellular PGN fragment, may act as a decoy to divert PGLYRP1 activity. However, no increase in the bactericidal activity of recombinant mPGLYRP1 was noted between WT or TCT-deficient strains following incubation with PGLYRP1 (Fig. 1H), arguing against this hypothesis. Given prior work showing that Bordetella polysaccharides confer resistance to antimicrobial peptides32, we tested whether extracellular polysaccharides inhibit PGLYRP1 function. The Bps operon is required for effective production of B. pertussis surface polysaccharide33 . Using a BpsB-deficient strain we observed significantly reduced resistance to mPGLYRP1 activity compared to WT bacteria (Fig. 1I, p< 0.01), suggesting that endogenous polysaccharides can shield B. pertussis from PGLYRP1 activity.

Together, these results indicate that PGLYRP1 plays a dual role during B. pertussis infection. PGLYRP1 enhances early bacterial control via neutrophil-mediated killing and direct bacteriostatic activity but may also play a deleterious role in controlling pulmonary infection at later stages of infection (Figure 1). While TCT does not appear to interfere with PGLYRP1 activity, Bordetella polysaccharides, previously associated with biofilm production33, do, highlighting a key bacterial evasion strategy against host innate defenses.

Peptidoglycan fragment TCT enhances PGLYRP1-mediated suppression of host responses in pertussis

Given the opposing effects of PGLYRP1 on bacterial burden at 4DPI and 7DPI, we hypothesized that PGLYRP1 exerts antibacterial activity at early time points but may suppress protective immune responses or enhance deleterious responses later in infection, impairing bacterial control by the host. Recently, PGLYRP1 has been associated with PGN recognition by innate immune receptors, including NOD2 and TREM-124, 26. However, B. pertussis releases a NOD1 activating PGN fragment, TCT14.

To assess the role of PGLYRP1 in B, pertussis-driven inflammation, we performed histological analysis of lungs from WT and PGLYRP1 KO mice at 4 and 7DPI. We observed significantly reduced immunopathology, based on the percentage and severity of bronchovascular bundle formation and infiltration of immune cells into alveolar spaces, in PGLYRP1 KO mice at both timepoints (Fig. 2A, p < 0.05 at 4DPI, p < 0.01 at 7DPI). These findings suggest that PGLYRP1 contributes to inflammatory lung damage during B. pertussis infection.

Assessment of PGLYRP1’s role in pulmonary inflammation and transcriptional responses during Bordetella pertussis infection.

Semi-quantitative scoring of hematoxylin and eosin (H&E) stained lung sections from wild-type BALB/c (black dots) and PGLYRP1 knockout (red dots) mice at 4- and 7-days post-infection with B. pertussis (A). Data represents average scores of 3 blinded investigators scoring degree and percentage of peribronchial infiltration and alveolar consolidation. Bulk RNA sequencing was performed on lung homogenates collected from WT and PGLYRP1 KO mice at 4DPI to assess differential gene expression. Volcano plots (B,D) and heat-maps (C,E) showing total (B,D) and immune-related (C,E) differentially expressed genes between B. pertussis challenged BALB/c (WT) and PGLYRP1 KO mice (B,C) or PGLYRP1 KO mice challenged with parental wild-type B. pertussis (WT) or TCT-under releasing strain TCT(-). (F) KEGG enrichment pathway analysis was performed on differentially expressed genes from lung tissue isolated from mice challenged with WT or TCT(-) B. pertussis. Data represent the mean of biological replicates; statistical analyses were performed by Mann Whitney U test p-values; *=0.003, **, p-value < 0.01

To investigate PGLYRP1s role in immune signaling during B. pertussis infection, we performed bulk RNA sequencing on lungs from infected BALB/c (WT) and PGLYRP1 KO mice. PGLYRP1 KO mice exhibited increased expression of pro-inflammatory cytokines (e.g., IL6, IL23A, IL1A, IL1B, and IL36G) and chemokines (e.g., CCL2, CCL3, CCL12, and CCL28) alongside reduced expression of anti-inflammatory cytokine IL10 (Fig. 2B-C). This transcriptomic profile suggests a model in which PGLYRP1 dampens key cytokine responses while paradoxically promoting immunopathology.

PGLYRP1 is a soluble PGN-binding protein. B. pertussis releases an extracellular PGN fragment, tracheal cytotoxin (TCT). We have recently demonstrated that TCT inhibits immune responses to B. pertussis18.We hypothesized that interactions with TCT contribute to PGLYRP1 influence on host inflammatory responses to B. pertussis. To test this, we compared the transcriptional responses of PGLYRP1 KO mice challenged with TCT-producing (parental) or TCT-deficient B. pertussis strains. Both strains induced a similar pattern of cytokine upregulation in the absence of PGLYRP1, but this increased inflammation was more robust in the absence of TCT (Fig. 2D&E). Specifically, PGLYRP1 KO mice infected with TCT(-) bacteria exhibited greater expression of IL23A, IL6, IL1B, and IL36G compared to WT infection, suggesting that TCT may potentiate the immune-dampening activity of PGLYRP1.

PGLYRP1 has previously been shown to stimulate inflammatory responses to PGN26. These findings indicate that the immunomodulatory function of PGLYRP1 is informed by the PGN it interacts with. TCT release may be a strategy employed by B. pertussis to co-opt PGLYRP1 function and suppress protective inflammation, thereby promoting pathogen persistence. Supporting this, KEGG pathway analysis on differential gene expression of 4DPI WT (parental) and TCT(-) lungs confirmed enrichment of cytokine-cytokine receptor interactions, NOD signaling, and IL-17/TNF pathways in the presence of TCT (Fig. 2F).

TCT-PGLYRP1 interactions drive NOD1 signaling

To understand how TCT modulates PGLYRP1-dependent immune signaling, we investigated known mechanisms by which PGLYRP1 influences host PRR signaling. Recent studies suggest that intracellular PGLYRP1 can act as a co-factor in NOD2 signaling24 and TCT is a known NOD1 agonist. Hence, we hypothesized that PGLYRP1 may enhance NOD1 signaling in response to the muropeptide TCT, a NOD1 exclusive agonist. To test this, recombinant PGLYRP1 was incubated with TCT (1 ng/mL) before stimulation of NOD1 and NOD2 reporter cells. Consistent with previous reports, TCT robustly stimulated a NOD1 response (Fig. 3A). Notably, both murine (Fig. 3A, p < 0.005) and human (Fig. 3B, p < 0.001) PGLYRP1 significantly enhanced TCT-mediated NOD1 activation compared to TCT alone. Importantly, PGLYRP1 alone did not induce NOD1 activation (Fig. 3A), suggesting the enhanced NOD1 stimulation was not by PGLYRP1 directly activating NOD1 (Fig. 3A). Interestingly, this enhanced activation was specific to NOD1 as neither TCT nor TCT+PGLYRP1 activated NOD2 reporter cells (data not shown) since TCT is not a NOD2 agonist. On the other hand, mPGLYRP1 decreased NOD2 recognition of NOD2 agonist muramyl dipeptdide (MDP, p < 0.01) (Fig. 3C), suggesting exogenous PGLYRP1 boosts responses to NOD1 stimulating PGN structures like TCT but dampens responses to NOD2 stimulatory PGNs.

PGLYRP1 selectively modulates NOD1 and NOD2 signaling in response to Bordetella pertussis peptidoglycan.

(A–B) HEK293 reporter cells expressing either mouse (A and D) or human (B) NOD1 or (C) NOD2 were stimulated with recombinant mouse (A) or murine or human (B-D) PGLYRP1 (12.5 ug/mL), purified tracheal cytotoxin (TCT, 1 ng/mL) (A,B), muramyl dipeptide (MDP, NOD2 agonist) (C), or conditioned bacterial growth media (OD = 0.8) (D). NF-κB-driven SEAP reporter activity was measured to assess pathway activation. Reporter activation was quantified after 18–24 hours and normalized to PBS-treated cell controls as a zero baseline. Single-cell RNA sequencing was performed on lung tissue from BALB/c mice infected with B. pertussis at 4 days post-infection. Neutrophils were subjected to further sub-clustering and visualized as a UMAP (E). Violin plots visualizing expression of Nod1, Nod2, IL1a and Pglyrp1 across neutrophil sub-clusters (F). Volcano plot demonstrating differentially expressed genes in Nod1 (blue) or Nod2 (red) expressing neutrophils (G) Single-cell analyses represent combined data from two mice per group. Statistical analyses were performed using Student’s t-test or adjusted p-values as appropriate. **, p-value < 0.01, *** p-value < 0.005, **** p-value < 0.001

To determine if PGLYRP1 modulates NOD responses to other potential PGNs released by B. pertussis, we stimulated reporter cells with mPGLYRP1 incubated with conditioned media from WT or TCT(-), or TCT(+) cultures. Consistent with our results using purified TCT, mPGLYRP1 amplified NOD1 signaling in response to B. pertussis conditioned media, likely due to increased recognition of extracellular TCT (Fig. 3D, p < 0.01). In contrast, NOD2 signaling was significantly suppressed by mPGLYRP1 (Fig. 3D), consistent with our results with purified PGN (Fig. 3A-C). Together, these findings suggest that exogenous PGLYRP1 acts as a selective modulator of NOD signaling, enhancing NOD1 activation while dampening NOD2 responses.

To assess the functional consequence of skewed NOD1 versus NOD2 expression and signaling during infection, we analyzed single-cell RNA sequencing data from lungs of BALB/c mice at 4DPI. We examined NOD1 and NOD2 expression across neutrophil sub-clusters (Fig. 3E&F), as neutrophils are the primary expressors of PGLYRP1 during B. pertussis infection (Fig. 1C). NOD2-expressing neutrophils showed higher expression of pro-inflammatory genes including Il1a, Ccl3 and Ptgs2 (COX-2) (Fig. 3F&G). In contrast, NOD1 expression correlated with genes involved in anti-inflammatory immune responses: Pla2g7 (degrades platelet activating factor), Tmsb4x which inhibits neutrophil migration, promotes repair and inhibits NFkB and increased expression of PGLYRP1, which may further positively regulate these pathways34. This identifies a correlation between pro-inflammatory cytokine expression and NOD2 signaling in neutrophils.

Interestingly, when we further analyzed the clusters with high Nod2 expression, like clusters 0 and 4, we observed gene expression patterns which correspond to previously identified inflammatory neutrophil populations. Cluster 0 hallmark genes were associated with interferon-stimulated programming (Calhm6, Cxcl9, Tma16, Slamf7, Baft2) (Sup. Fig. 1), and B. pertussis infection has been shown to cause both deleterious and beneficial interferon responses34-36. Additionally, poor B. pertussis host-responses have been linked to interferon-stimulated neutrophils, providing interesting context to this identified neutrophil cluster37, 38. Furthermore, cluster 4 marker (Hilpda, Csf3, Myl4, Odc1, Spp1) neutrophils have been identified in tumor exhaustion environments, and related to hypoxia response, highlighting another inflammatory neutrophil population (Sup. Fig. 1)39.

These results highlight the heightened inflammatory phenotype of NOD2+ neutrophils and suggest that TCT-mediated skewing of NOD responses towards NOD1, amplified by PGLYRP1, may be a mechanism to suppress inflammation and promote B. pertussis persistence.

TCT-PGLYRP1 complexes dampen TREM-1 activation, limiting inflammation during B. pertussis infection

In addition to their role in NOD signaling, PGLYRP1-PGN complexes have been shown to activate TREM-1, driving NFkB-mediated inflammatory responses26. This ability to activate TREM-1 is boosted by the presence of a PGN from the sacculi of S. aureus, B. subtilis or E. coli26. To determine if TREM-1 contributes to host responses to B. pertussis, we first assessed TREM-1 expressing cell-types following B. pertussis infection in mice using single cell RNA-sequencing. Following B. pertussis infection, TREM-1 was primarily expressed by neutrophils, with lesser expression noted in monocytes and macrophages (Fig. 4A).

TCT inhibits PGLYRP1 signaling via TREM-1 Violin plot generated from single-cell RNA sequencing (scRNA-seq) displaying TREM-1 expression on neutrophils (green), alveolar macrophages (AM, gold), and inflammatory monocytes (IM) at 4DP1 (A).

TREM-1 activation measured via NFAT-driven luminescence (relative luminescence units, RLU) in TREM-1 reporter cells treated with recombinant human PGLYRP1 with or without 5µg/mL of Lys-type peptidoglycan derived from Staphylococcus aureus (5 µg/mL), TCT (5 µg/mL) or a DAP-type PGN from Bacillus subtilis (5 µg/mL) (B). Diagram depicting proposed model (C). All TREM-1 activation assays were performed with biological triplicates (n = 3). Statistical analyses were performed using unpaired two-tailed Student’s t-test. *, p-value < 0.05, **, p-value < 0.01, *** p-value < 0.005, **** p-value < 0.001

To determine if TCT influences PGLYRP1-dependent TREM-1 activation, we assessed the ability of TCT, a DAP-type PGN monomer, to enhance PGLYRP1-mediated TREM-1 activation. As reported previously, preparations of peptidoglycans isolated from S. aureus and B. subtilis enhanced PGLYRP1-mediated activation of a TREM-1 reporter cell line (Fig. 4B, p < 0.01)40, 41. Given the anti-inflammatory role of TCT observed in our data (18 and Figs 2 and 3), and known role for TREM-1 in B. pertussis pathogenesis42 we proposed that TCT may be a poorer enhancer of PGLYRP-1 mediated TREM-1 activation. Consistent with our hypothesis the addition of TCT significantly reduced TREM-1 activity (Fig. 4C). This suggests that TCT, unlike PGN from the sacculi, fails to support and possibly even antagonizes PGLYRP1’s ability to activate TREM-1 (Fig. 4C). Further studies are required to discern the contribution of PGN structure and solubility in this model.. Combined with our previous work42, these findings highlight the importance of PGN in bacterial regulation of TREM-1 and its role in B. pertussis pathogenesis and disease and the potential impact of skewing TREM-1 responses.

Discussion

Our study reveals a previously unrecognized role for the PRR, PGLYRP1, in orchestrating both antibacterial defenses and inflammatory responses during Bordetella pertussis infection. While traditionally characterized as a bactericidal effector which mediates its effects by binding PGN on the surface of Gram-positive bacteria, emerging work suggests its role may extend to mammalian immune signaling25. The contribution of PGLYRP1 to PRR mediated inflammatory responses in mammals and in particular how PGN structure impacts the outcomes of these interactions, remains largely unexplored. Our findings demonstrate that PGLYRP1 can functions as a PGN structure-specific interpreter which translates structure into context and pathogen specific immune signaling. This dual role expands our understanding of how innate immune receptors fine-tune host-pathogen interactions in the airway based on the structure of PGN encountered. Prior studies have also shown that endogenous PGLYRP1 bolstered NOD2-driven macrophage activation in response to GMTriP-K, but not in response to related PGN fragments such as GMDiP or MDP, highlighting the structural specificity of PGLYRP1-mediated immune recognition24. We expand on these findings to suggest exogenous PGLYRP1 biases NOD1 responses and dampens NOD2 responses. Furthermore, we suggest pathogens like B. pertussis have evolved to release PGN structures, such as TCT, which dampen the ability of PGLYRP1 to signal via TREM-1 (Fig. 4C), suggesting an active area of ongoing host-pathogen evolution.

Here, we propose PGLYRP1 as a PGN-structure specific regulator of inflammatory responses. This is comparable to SIGLECs, which can distinguish self from non-self glycans to ignore or elicit immune responses43. During B. pertussis infection, PGLYRP1, predominantly expressed by neutrophils, contributes to early bacterial clearance, as PGLYRP1 KO mice exhibit significantly higher bacterial loads at 4DPI. The bactericidal activity of PGLYRP1 is diminished by the presence of extracellular polysaccharide, highlighting the potential for Bordetella pertussis-derived extracellular decoys to prevent antibiotic activity. PGLYRP1 KO mice exhibit elevated inflammatory cytokine expression, suggesting PGLYRP1 also serves an anti-inflammatory function. Transcriptomic analysis of PGLYRP1 KO lungs infected with mutants which fail to release TCT or the parental strain support the hypothesis that PGLYRP1 limits inflammation in a PGN structure-dependent manner: specifically, the monomeric PGN fragment TCT potentiates the anti-inflammatory activity of PGLYRP1. An alternative explanation for the discrepancy observed in bacterial burden at early and later timepoints of infection is that PGLYRP1-directed lysis of B. pertussis releases more host-available muropeptides, which PGLYRP1 can then interact with to enhance less protective NOD1-driven immune responses at the expense of more potent NOD2-driven responses (Fig 1)18.

Mechanistically, PGLYRP1 enhances NOD1 signaling in response to TCT while suppressing NOD2 signaling in response to MDP directing the polarization of immune responses away from a more inflammatory NOD2-driven response. Single cell RNA sequencing revealed that PGLYRP1 expression is enriched in NOD1+ neutrophils, which have less inflammatory profiles than NOD2+ neutrophils. These data suggest B. pertussis may exploit PGLYRP1 to amplify a NOD1-dominant signaling environment that blunts host inflammation. Recent work from our group highlighted the ability of TCT to polarize NOD responses, towards NOD1 as an immune evasion mechanism18. This work expands on this finding, highlighting the role of PGLYRP1 and PGN structure in these findings, but further work should validate and expand on how PGLYRP1 modulates responses to muropeptide outside the cellular and murine models used in these studies.

These findings shift our understanding of mammalian PGLYRP1 from a passive bactericidal protein to an immunomodulator, whose contribution to inflammation is determined by the structure of PGN it encounters. Additionally, they inform our views of PGN structure and how it impacts host responses, suggesting bacterial release of PGNs, as a mechanism to dampen host immunity. These results propose a novel paradigm in host-pathogen interactions; that bacteria may release structurally tailored PGN fragments to bias host immune sensors and signaling in their favor.

Excitingly, we report a novel axis by which PGLYRP1 regulates TREM-1, a known amplifier of inflammation. TREM-1 activation has been reported to strengthen NF-kB signaling pathways which NLRs like NOD1 and NOD2 canonically signal through27. While PGLYRP1 synergized with PGNs from S. aureus and a B. subtilis to enhance TREM-1 activation, the presence of TCT antagonized this effect (Fig. 4). There are several differences in the structures of these PGNs: unlike monomeric TCT, PGN tested from other bacteria are derived from the sacculus and are insoluble, additionally B. subtilis PGN is amidated. Further studies are required to discern if solubility, size or other factors are responsible for differences in the effect on PGLYRP1-mediated TREM-1 activation. Additionally, TCT contains a 1,6 anhydroMurNAc. Bifidobacterium have been predicted to produce abundant anhydro-PGNs which markedly reduce LPS-stimulated cytokine expression in RAW264.7 cells, underscoring a potential anti-inflammatory effect of 1,6-anhydro-MurNAc containing PGNs, such as TCT.

Previous work from our group demonstrated that inhibition of TREM-1 prevented immunopathology and inhibited cytokine and chemokine responses to infection, highlighting the potential utility of disturbed TREM-1 signaling for B. pertussis42. TREM-1 inhibitor treated mice showed reduced inflammation and inflammatory cytokine responses following B. pertussis infection highlighting the importance of this pathway in pertussis pathogenesis and the potential utility of its manipulation by bacteria. Further, our findings suggest that selectively targeting TREM-1 or PGN structure sensing by PGLYRP1 could attenuate disease severity without compromising bacterial clearance, a strategy aligned with emerging host-directed therapeutics in TB and sepsis44-46. However, the role of these players in generating effective antibody responses still needs to be elucidated so that we can understand the implications on long-term immunity.

We hypothesize that B. pertussis may attenuate host immune responses not merely by evading recognition, but by actively hijacking host regulatory circuits through the release of specific PGN structures. Taken together, our findings support a model in which PGLYRP1 serves dual and temporally distinct roles during infection, by promoting bacterial killing while restraining inflammation in a PGN-structure specific manner. This duality may reflect temporal or disease stage specific changes in the types of PGN available to host PRRs, as 1,6-anhydro MurNAc containing PGNs are usually internal, unless released following bacterial lysis. The ability of B. pertussis to exploit this shift by releasing TCT to suppress excessive inflammation represents a novel immune evasion strategy.

More broadly, this work demonstrates that the structure of PGN fragments is a key determinant of host immune outcomes. Innate receptors like PGLYRP1 relay this structural information to shape both defense and pathology, providing an avenue for pathogen hijacking by alternating the availability of different PGN structures. Targeting the TCT–PGLYRP1–TREM-1 axis may offer a novel therapeutic or vaccine-mediated approach to mitigate immunopathology in pertussis and potentially other infections involving secreted PGN. These findings reframe PGN as both a pathogen-associated molecular pattern and an immunomodulatory cue that both host and pathogen exploit to regulate the immune-inflammatory environment. Future studies using defined synthetic muropeptides will be necessary to pinpoint structural determinants of the reduced immune activation noted toward TCT. Likewise, how these alterations in PGN structure impact PGLYRP1 interactions with host PRRs remains unclear.

Methods and materials

Mice and Ethics Statement

All animal procedures were approved by the Institutional Animal Care and Use Committee at the University of Maryland School of Medicine (protocol 00000108). Adult (6–8-week-old) BALB/c (Charles River, 028) and PGLYRP1 knockout (Pglyrp1tm1Rdz/Pglyrp1tm1Rdz in BALB/c background, generously provided by Dr. Roman Dziarski47), mice were bred and housed under specific pathogen-free conditions. To control for and assess the potential of sex-related differences in pertussis pathogenesis both male and female mice were used in experiments. For transcriptomic studies, RNA was isolated from the lungs of PBS- or B. pertussis challenged female mice.

Bacterial Strains and Infections

B. pertussis strains, including wild-type BC36, a streptomycin-resistant Tohama I derivative, or isogenic mutants (TCT-, TCT+, BpsB KO) were cultured on Bordet-Gengou agar supplemented with 10% defibrinated sheep blood for in vivo infections or in Stainer-Scholte (SS) broth for in vitro assays. Strains were grown in SS broth to mid-log stage (OD = 0.8), and the resulting media was centrifuged and filtered to remove any bacteria to make the conditioned media used for cell culture assays. TCT mutants were generously supplied by William Goldman (University of North Carolina at Chapel Hill). TCT-strain was generated by expression of the E. coli AmpG in BC36, resulting in a 50-fold decrease in TCT release. TCT+ was generated by replacing native B. pertussis AmpG with a kanamycin resistance cassette, this resulted in a 24-fold increase in TCT release30. Adult mice were anesthetized and inoculated intranasally with 2 × 106 CFU of B. pertussis in 50 µL PBS. Lungs were harvested at 4DPI and 7DPI for bacterial burden analysis by plating of serially-diluted lung homogenates.

RNA isolation and processing

Lung tissue was harvested at 2 or 7DPI for RNA isolation by the Trizol-choloroform method. Briefly, lung tissue was homogenized using the Omni Bead Beader (Omni, Inc), phase separated using chloroform, and precipitated using isopropanol. RNA was then quantified and converted to cDNA (iScript cDNA Synthesis Kit). Quantitative real-time PCR (qPCR) was performed using the BIORAD CFX96 real-time PCR instrument. Gene expression was calculated as fold change relative to PBS-inoculated control animals using the 2^−ΔΔCT threshold cycle method and normalized to Hypoxanthine phosphoribosyltransferase (HPRT) (internal housekeeping gene).

Bulk RNA Sequencing

At 4DPI, lungs were harvested from 3 WT or PGLYRP1 KO BALB/c mice infected with WT or TCT-deficient B. pertussis. Total RNA was extracted from homogenized lungs using Trizol Reagent (Invitrogen) according to the manufacturer’s protocol. RNA integrity was confirmed using a Bioanalyzer (Agilent), alignments were generated by HISAT2 sequencing performed using a Illumina NextSeq 1000. FastQC and RSeQC were used to assess read and alignment quality. Samtools was used to generate alignment statistics. Reads were aligned with the Mus musculus GRCm39 reference genome. Differential gene expression was analyzed using DESeq2, with significance defined as adjusted p-value < 0.05.

Single-Cell RNA Sequencing

Single-cell suspensions were prepared from lung tissues of duplicate infected and uninfected control C57BL/6 mice at 4DPI, barcoded and sequenced as individual mice. Following red blood cell lysis, cell viability was assessed by trypan blue staining, viable cells were sorted and loaded onto the 10x Genomics Chromium Controller for barcoding and library construction. Sequencing was performed on an Illumina NovaSeq platform. Data were processed using Cell Ranger and analyzed using the Seurat package in R. Quality control cutoffs of between 1000 and 10000 features and <20% mitochondrial DNA. Data were normalized and scaled in Seurat v5, and groups were integrated via Harmony-based integration. Clusters were identified, determined, and visualized via Seurat library workflow (RunPCA, FindNeighbors, FindClusters, RunUMAP) and then annotated via hallmark markers determined by differential gene expression analysis across clusters using Wilcoxon rank-sum tests. This dataset is now deposited with the Gene Expression Omnibus (GSE324217)

Neutrophil Isolation and Bactericidal Assays

Bone marrow-derived neutrophils (BMDNs) were isolated by density gradient centrifugation from adult C57BL6 mice and cultured in DMEM supplemented with 10% FBS. For killing assays, BMDNs were incubated with B. pertussis at a multiplicity of infection (MOI) of 10:1 for 2 or 24 hours. Supernatants were serially diluted and plated to quantify surviving CFU. Recombinant murine PGLYRP1 (R&D Systems) or BSA control was incubated with live B. pertussis or purified PGN fragments (TCT gifted by Prof. William Goldman or S. aureus PGN, Invivogen) to assess direct bactericidal activity.

RNAscope In Situ Hybridization

RNA in situ hybridization was performed on 5um formalin fixed paraffin embedded lung sections with the RNAscope detection kit to confirm RNA sequencing identification of PGLYRP1 transcripts (Advanced Cell Diagnostics). Hybridization, amplification, and detection steps were carried out according to the manufacturer’s protocol. Briefly, slides were deparaffinized in xylene and rehydrated using an ethanol series. Sections were treated with hydrogen peroxide for 10 minutes at room temperature, followed by treatment with target retrieval reagent (Advanced Cell Diagnostics). Sections were hybridized with target-specific probes (Advanced Cell Diagnostics) followed by amplification and fluorescent detection (Opal 570 and Opal 520). Slides were counterstained with hematoxylin and eosin and imaged using an ECHO Revolve 4 and analyzed with ImageJ.

Reporter Cell Assays for NOD1, NOD2, and TREM-1 Activation

HEK293-derived reporter cells expressing murine or human NOD1 and NOD2 with an, or TREM-1 and an NFkB–dependent luciferase or NF-kB inducible SEAP reporter (InvivoGen) and TREM-1-DAP12 reporter cells (Eurofins Discover X )in the Jurkat T cell background with an NFAT-dependent luciferase were stimulated with B. pertussis tracheal cytotoxin TCT (1 ng/mL), conditioned media (1:20 ratio of conditioned bacterial media to cell culture media), PGN fragments (TCT, S. aureus, or B. subtilis PGN (Invivogen)) in the presence or absence of recombinant human PGLYRP1 (R&D systems) at a concentration of 5ug per well for 24 hours. Supernatants were analyzed for reporter activity using colorimetric (NOD1 and NOD2 reporter cells; Promega) or luminescent (TREM-1 reporter cells; Invivogen) detection kits (Promega). TCT was quantitated from its peak area elution profile from the final purification step, For each preparation of TCT this area was compared to the peak area of a purified standard that had been quantitated by amino acid analysis and generously provided by William Goldman, (UNC)

Histopathology

Lungs were fixed in 10% formalin, embedded in paraffin, and sectioned at 5 µm thickness. Slides were stained with hematoxylin and eosin (H&E). Scores are given as the average score from 3 blinded investigators based on percentage and severity of bronchovascular bundle formation and infiltration of immune cells into alveolar spaces.

Statistics

Significant differences in pathology scores were calculated at each timepoint using a 2-tailed Mann-Whitney U test. Each timepoint was analyzed independently. A p-value < 0.05 was considered significant. A p-value < 0.05 is denoted by *, < 0.01 by ** and < 0.005 by ***